Water Quality at Toronto Beaches

### 2024-09-26 paper/paper.qmd
Paper draft 3 (adds references also)

---
title: "Water Quality at Toronto Beaches"
subtitle: "An analysis of e-coli rates in toronto beaches 2007-2024"
author: 
  - Jacob Gilbert
thanks: "Code and data are available at: https://github.com/JfpGilbert0/toronto-beaches-water-quality."
date: today
date-format: long
abstract: ""
format: pdf
jupyter: python3
---

```{python}
#| echo: false
# import dependancies
import jupyter
```

# Introduction
Water contamination is important in Toronto, especially in the summer months as locals flock to the nearby beaches as the weather turns from snow to sun. E-coli testing is common at most public swimming locations as high levels highlight risk to those embarking into the water. While E-coli is unlikely to cause illness on it's own, If a water sample has high concentrations of E. coli, other more dangerous and infectious organisms may be present.(https://www.cleanlakesalliance.org/e-coli/). Thus the levels at public beaches can affect public health, and as such action should be taken on the results. To either close public beaches that exhibit unsafe levels, or continuation of previous action that has led to lower levels. 
Despite these public health measures, there is limited understanding of the specific patterns of E. coli contamination in Toronto's beaches. Most public advisories and beach management decisions rely on reactive measures rather than proactive analysis of historical trends. This paper aims to fill this gap by examining historical E. coli data from Open Data Toronto, focusing on two of the city’s popular beaches: Sunnyside Beach and Marie Curtis Park East Beach. By analyzing data over several years, this study seeks to identify temporal patterns in contamination, such as variations by day of the week and season, to help answer the critical question: When are beachgoers at the highest risk of swimming in contaminated waters?
Additionally, this analysis explores whether E. coli levels have shown improvement over time, informing on the effectiveness of current water management practices. Understanding the current and historic trends are paramount for city officials to understand if current water quality measures are sufficient. Or in the worse case, that interventions are needed in order to keep Toronto beaches safe for everyone.
This paper is structured as follows: First, we present the dataset, describing its sources, key variables, and the data cleaning process. Next, we analyse patterns in E. coli levels, examining how contamination varies across different times and locations. We then discuss the implications of these findings for public health and beach management, followed by a conclusion that summarises the key insights and recommendations for future action.

## Referencing
In this paper, Python [] and it packages are utilized as the primary programming language for the following analysis and displaying the data. Key packages used include pandas [McKinney, 2010], which facilitated efficient data cleaning, manipulation, and summarization of the dataset, NumPy [Harris et al., 2020] was employed for numerical operations, Matplotlib [Hunter, 2007] and seaborn [Waskom, 2021] were critical for creating informative visualizations.

@@

## Data Source
Opendatatoronto is a publicly available collection of wide ranges of data. This resource was used to obtain the e-coli levels of the water at two of Toronto's many beaches. The data set included coli levels from the SunnySide and Marie Curtis park east beaches, from the 3rd of July 2007 to September 5th 2024. The large date range allows us to extrapolate trends well and further the recency of the data allows us to generate a good understanding of the present. We have a total of 3084 data points from the two beaches, 1370 and 1714 data points respectively. Each detailing the collectiondate, beachname, sitename, geometric location (given by longitude and latitude) and the e coli levels.

@@

## Measurement
E-coli is measured in MLN/100ml a measurement that reflects the ecoli particles in evey 100ml of lake water. For our data presentation log values are used to give a better visualisation of the data. To obtain an accurate representation of the beach as a whole 12 different locations here used to gather data across the 2 beaches as to remove bias that one testing location may result in. the result is that we obtain an unbiased representation of ecoli levels at each beach as a whole over more than a decade. Some missing data we encounter however is that although we have the collection date we do not observe the time that data was collected. Missing this data may lead to some bias as the time of day could impact the baseline levels of e-coli in the water. What this form of measurement does allow us to do is look at the differences in water quality at a date level, a figure() shows there is a good distribution of data from each day of the week.
Ther are some gaps in the data however, in ehe monthly data we only find rsults from the summer months. As during the summer the impact of water quality is at its highest this doesn't impact the usefulnss of the data, howevr it does result in missing context of levels during the rest of the year.

# Methods
Before we look at the daily differences in levels we will look at the data as a whole, observing the distribution of higher and lower levels of ecoli in the water across time as well as by beach.  Using canada.ca we observe that they define unsafe levels of ecoli in drinking water to ber 88ML/100ml. As such we will use this in our analysis to observe the frequency that “unsafe” levels are observed in the beach water. Finally we will look at these distributions over the course of a week to answer the question of when water quality is at its best, and as a lake swimmer when is it safest for me to swim at these beaches.

# Results
## Time Series
```{python}
#| echo: false
# Execute the external Python script
with open("../code/model_generation/ecoli_scatter.py") as script_file:
    exec(script_file.read())
```

@@

Figure 1 is a scatter plot of E. coli levels over time by site and beach, and displays the data as a whole. It highlights the trends in contamination levels at Sunnyside Beach and Marie Curtis Park East Beach from 2008 to 2024.the beach site itself does not seem to significantly impact the overall trend of E. coli levels over time, as both Sunnyside and Marie Curtis Park East show similar patterns of fluctuation. the magnitude of the fluctuations ar very notable as prior to 2017 there are spikes in e-coli levels to many degrees, some extremely high. Whereas afer 2017 we see levels almost exclusivly sitting at 10 or around 1000 MNP/100ml. This appears to indicate that although the risks are persistant over time they are lessend in recent years.

Figure 2 is a bar chart illustrating the percentage of tests above the unsafe threshold by year, provides an important contrast to the scatter plot. Although the scatter plot suggests a reduction in extreme E. coli levels in the later years, this bar graph reveals that the frequency of tests exceeding the 100 MPN/100mL threshold does not diminish in the same way. Years such as 2010, 2011, 2019, and 2021 see substantial spikes in the percentage of unsafe tests, reaching as high as over 50% in 2010. We do not see percentages moving consitently in on direction other than after 2017 where the rates appear to be climbing. 2017 is a strang anomoly here with no tests returning unsafe results. The combination of these two graphs paint a clear picture of water quality in Toronto over the years.

```{python}
#| echo: false
#| label: fig-2
# Execute the external Python script for individual beach scatter plots
with open("../code/model_generation/yearly_percentages.py") as script_file:
    exec(script_file.read())
```

**Figure 2**: Bar graph reflecting the percentage of results each year where the water quality exceeds 100MPN/100ml. There is no obvious pattern here although in later years we see a trend upwards and rates appear to trend down until 2017 where the rate is 0.

## Monthly data
```{python}
#| echo: false
#| label: fig-3
# Execute the external Python script for monthly summary statistics
with open("../code/model_generation/monthly_summary.py") as script_file:
    exec(script_file.read())
summary
```

@@

Figure 4 looks at the monthly data at each beach, as mentioned in the data section only th summer months provide data due to the cold weather in toronto limiting available data in the winter. Further we see a disproportionate number of daa points in May and September, this could make the results here less reliable. However this data can still provide useful data. 
May exhibits the lowest average E. coli levels, at 101.76 MPN/100mL, which is just above the unsafe threshold. With only 6.16% of samples being unsafe. The low variance suggests that water quality is fairly consistent during this early summer month. 
August and July show extremely high averages, above 500MPN/100ml, with over 30% of daa collected here showing water quality above acceptable levels. Looking at figure() these high levels are consistent between both beaches, though we do observe extremely high average levels at Sunnyside.
Overall, we see the average level aswell a the likelyhood of unsafe levels increase as the summer progresses and slighly taper off in september. Although the absence of a reasonable amount of data in September could contribute this increasing trend.

```{python}
#| echo: false
# Execute the external Python script
with open(f"../code/model_generation/monthly_ecoli.py") as script_file:
    exec(script_file.read())
```

## Weekly Data
```{python}
#| echo: false
#| label: fig-5
# Execute the external Python script for weekly summary statistics
with open("../code/model_generation/weekly_summary.py") as script_file:
    exec(script_file.read())
summary
```

@@

```{python}
#| echo: false
#| label: fig-6
# Execute the external Python script for weekly E. coli levels by beach
with open("../code/model_generation/dow_bar.py") as script_file:
    exec(script_file.read())
```

**Figure 6**: E. coli levels by day of the week show significant differences between the two beaches, with Sunnyside exhibiting extraordinarily high levels on Monday and Tuesday compared to Marie Curtis.

# Conclusion
This analysis of E. coli contamination levels at Sunnyside Beach and Marie Curtis Park East Beach reveals critical insights into the patterns and trends of water quality in Toronto’s public swimming areas. While the data shows a reduction in the variability of E. coli levels in recent years, the frequency of unsafe readings remains concerning, particularly during peak summer months. The persistent risk highlighted by the bar graph indicates that ongoing contamination remains a public health challenge, despite efforts to improve water quality. Policy makers should improve their effors to reduce the number of days that water quality is unacceptable over the summer months.
The months that are exhibiting the greates risk are..

>>>>

---
title: "Water Quality at Toronto Beaches"
subtitle: "An analysis of e-coli rates in toronto beaches 2007-2024"
author:
  - Jacob Gilbert
thanks: "Code and data are available at: https://github.com/JfpGilbert0/toronto-beaches-water-quality."
date: today
date-format: long
abstract: "This study analyses E. coli contamination trends at two of Toronto's popular beaches, Sunnyside Beach and Marie Curtis Park East Beach, from 2007 to 2024. Using data sourced from Open Data Toronto, the analysis examines temporal patterns in contamination levels, assessing the frequency and severity of unsafe E. coli concentrations as defined by public health thresholds. The study reveals that while E. coli levels appear less erratic and generally lower in magnitude in recent years, the frequency of tests exceeding unsafe thresholds has not significantly diminished, indicating persistent contamination risks. These findings underscore the need for continued monitoring and targeted interventions to protect public health. This research aims to inform beach management practices and guide future efforts to improve water quality at Toronto’s beaches, Aswell as provide insights for swimmers on the risks that they are exposed to."
format: pdf
jupyter: python3
bibliography: references.bib
---

```{python}
#| echo: false
# import dependencies
import jupyter
```

# Introduction
Water contamination is important in Toronto, especially in the summer months as locals flock to the nearby beaches as the weather turns from snow to sun. E-coli testing is common at most public swimming locations as high levels highlight risk to those embarking into the water. While E-coli is unlikely to cause illness on its own, If a water sample has high concentrations of E. coli, other more dangerous and infectious organisms may be present [@cleanlakesalliance]. Thus the levels at public beaches can affect public health, and as such action should be taken on the results. To either close public beaches that exhibit unsafe levels, or continuation of previous action that has led to lower levels.
Despite these public health measures, there is limited understanding of the specific patterns of E. coli contamination in Toronto's beaches. Most public advisories and beach management decisions rely on reactive measures rather than proactive analysis of historical trends. This paper aims to fill this gap by examining historical E. coli data from Open Data Toronto, focusing on two of the city’s popular beaches: Sunnyside Beach and Marie Curtis Park East Beach. By analysing data over several years, this study seeks to identify temporal patterns in contamination, such as variations by day of the week and season, to help answer the critical question: When are beachgoers at the highest risk of swimming in contaminated waters?
Additionally, this analysis explores whether E. coli levels have shown improvement over time, informing on the effectiveness of current water management practices. Understanding the current and historic trends are paramount for city officials to understand if current water quality measures are sufficient. Or in the worse case, that interventions are needed in order to keep Toronto beaches safe for everyone.
This paper is structured as follows: First, we present the dataset, describing its sources, key variables, and the data cleaning process. Next, we analyse patterns in E. coli levels, examining how contamination varies across different times and locations. We then discuss the implications of these findings for public health and beach management, followed by a conclusion that summarises the key insights and recommendations for future action.

## Referencing
In this paper, Python [@python] and its packages are utilised as the primary programming language for the following analysis and displaying the data. Key packages used include pandas [McKinney, 2010], which facilitated efficient data cleaning, manipulation, and summarization of the dataset, NumPy [@numpy] was employed for numerical operations, Matplotlib [matplotlib] and seaborn [@seaborn] were critical for creating informative visualisations.

@@

## Data Source
Opendatatoronto is a publicly available collection of wide ranges of data. This resource was used to obtain the e-coli levels of the water at two of Toronto's many beaches. The data set included coli levels from the SunnySide and Marie Curtis park east beaches, from the 3rd of July 2007 to September 5th 2024. The large data range allows us to extrapolate trends well and further the recency of the data allows us to generate a good understanding of the present. We have a total of 3084 data points from the two beaches, 1370 and 1714 data points respectively. Each detailing the collectiondate, beachname, sitename, geometric location (given by longitude and latitude) and the e coli levels.

@@

## Measurement
E-coli is measured in MPN/100ml a measurement that reflects the e coli particles in every 100ml of lake water. For our data presentation log values are used to give a better visualisation of the data. To obtain an accurate representation of the beach as a whole 12 different locations here used to gather data across the 2 beaches as to remove bias that one testing location may result in. The result is that we obtain an unbiased representation of e coli levels at each beach as a whole over more than a decade. Some missing data we encounter however is that although we have the collection date we do not observe the time that data was collected. Missing this data may lead to some bias as the time of day could impact the baseline levels of e-coli in the water. What this form of measurement does allow us to do is look at the differences in water quality at a date level, a figure() shows there is a good distribution of data from each day of the week.
There are some gaps in the data however, in the monthly data we only find results from the summer months. As during the summer the impact of water quality is at its highest this doesn't impact the usefulness of the data, however it does result in missing context of levels during the rest of the year.

# Methods
Before we look at the daily differences in levels we will look at the data as a whole, observing the distribution of higher and lower levels of e coli in the water across time as well as by beach. The Ontario government sets their standard of 100 mpn/100ml as their water quality objectives [@ontariowater]. Thus we will use this as our benchmark for safe e-coli levels in the Beach water. Finally we will look at these distributions over the course of a week to answer the question of when water quality is at its best, and as a lake swimmer when is it safest for me to swim at these beaches.

# Results
## Time Series
```{python}
#| echo: false
#| label: fig-1
# Execute the external Python script
with open("../code/model_generation/ecoli_scatter.py") as script_file:
    exec(script_file.read())
```
**Figure 1** depicts E. coli contamination levels over time at Sunnyside Beach and Marie Curtis Park East Beach from 2008 to 2024. The scatter plot shows that E. coli levels were highly variable and frequently above the unsafe threshold in the early years, especially between 2008 and 2013. From 2017 onwards, E. coli levels appear less erratic, with fewer instances of extreme spikes

@@

Figure 1 is a scatter plot of E. coli levels over time by site and beach, and displays the data as a whole. It highlights the trends in contamination levels at Sunnyside Beach and Marie Curtis Park East Beach from 2008 to 2024.the beach site itself does not seem to significantly impact the overall trend of E. coli levels over time, as both Sunnyside and Marie Curtis Park East show similar patterns of fluctuation. The magnitude of the fluctuations are very notable as prior to 2017 there are spikes in e-coli levels to many degrees, some extremely high. Whereas after 2017 we see levels almost exclusively sitting at 10 or around 1000 MNP/100ml. This appears to indicate that although the risks are persistent over time they are lessened in recent years.

Figure 2 is a bar chart illustrating the percentage of tests above the unsafe threshold by year, providing an important contrast to the scatter plot. Although the scatter plot suggests a reduction in extreme E. coli levels in the later years, this bar graph reveals that the frequency of tests exceeding the 100 MPN/100mL threshold does not diminish in the same way. Years such as 2010, 2011, 2019, and 2021 see substantial spikes in the percentage of unsafe tests, reaching as high as over 50% in 2010. We do not see percentages moving consistently in one direction other than after 2017 where the rates appear to be climbing. 2017 is a strange anomaly here with no tests returning unsafe results. The combination of these two graphs paint a clear picture of water quality in Toronto over the years.

```{python}
#| echo: false
#| label: fig-2
with open("../code/model_generation/yearly_percentages.py") as script_file:
    exec(script_file.read())
```

**Figure 2**: Bar graph reflecting the percentage of results each year where the water quality exceeds 100 MPN/100ml. There is no obvious pattern here although in later years we see a trend upwards and rates appear to trend down until 2017 where the rate is 0.

## Monthly data
```{python}
#| echo: false
#| fig.pos: "H"
#| label: fig-3
# Execute the external Python script for monthly summary statistics
with open("../code/model_generation/monthly_summary.py") as script_file:
    exec(script_file.read())
summary
```

@@

Figure 4 looks at the monthly data at each beach, as mentioned in the data section only the summer months provide data due to the cold weather in Toronto limiting available data in the winter. Further we see a disproportionate number of data points in May and September, this could make the results here less reliable. However this data can still provide useful data.
May exhibits the lowest average E. coli levels, at 101.76 MPN/100mL, which is just above the unsafe threshold. With only 6.16% of samples being unsafe. The low variance suggests that water quality is fairly consistent during this early summer month.
August and July show extremely high averages, above 500 MPN/100ml, with over 30% of data collected here showing water quality above acceptable levels. Looking at figure() these high levels are consistent between both beaches, though we do observe extremely high average levels at Sunnyside.
Overall, we see the average level as well as the likelihood of unsafe levels increase as the summer progresses and slightly taper off in september. Although the absence of a reasonable amount of data in September could contribute this increasing trend.

```{python}
#| echo: false
#| label: fig-4
#| fig.pos: "H"
# Execute the external Python script
with open(f"../code/model_generation/monthly_ecoli.py") as script_file:
    exec(script_file.read())
```

**figure 4** A bar graph showing the percent of results that are above the government target of 100 mpn/100ml grouped by month. The rates peak at both beaches in the middle months. Sunnysde experienced the highest rates of the two locations but also the lowest. Marie curtis had a bell-curve type of trend over the 4 summer months peaking in July.

\newpage
## Weekly Data
```{python}
#| echo: false
#| fig.pos: "H"
#| label: fig-5
# Execute the external Python script for weekly summary statistics
with open("../code/model_generation/weekly_summary.py") as script_file:
    exec(script_file.read())
summary
```

@@

\newpage
```{python}
#| echo: false
#| fig.pos: "H"
#| label: fig-6
# Execute the external Python script for weekly E. coli levels by beach
with open("../code/model_generation/dow_bar.py") as script_file:
    exec(script_file.read())
```
**Figure 6**: E. coli levels by day of the week show significant differences between the two beaches, with Sunnyside exhibiting extraordinarily high levels on Monday and Tuesday compared to Marie Curtis.

# Conclusion
This analysis of E. coli contamination levels at Sunnyside Beach and Marie Curtis Park East Beach reveals critical insights into the patterns and trends of water quality in Toronto’s public swimming areas. While the data shows a reduction in the variability of E. coli levels in recent years, the frequency of unsafe readings remains concerning, particularly during peak summer months. The persistent risk highlighted by the bar graph indicates that ongoing contamination remains a public health challenge, despite efforts to improve water quality. Policy makers should improve their efforts to reduce the number of days that water quality is unacceptable over the summer months.
The months that are exhibiting the greatest risk are July and August, as such Policy makers should focus their efforts on this time of the year. As these months fall in the middle of the summer the higher levels could be a result of climate or increased usage of the beaches in the warmer weather.
From the perspective of swimmers the results displayed here give insights on how to lower the risks of visiting the beaches when water quality is lower. Firstly overall we see that in recent years the extremes are much lower which is good news for lake swimmers in toronto. We also see that choosing to swim in the earlier months of the summer appear to bring with them better overall water quality than later months. Similarly the weekends bear a far lower risk to swimmers, which is fortunate as it is likely that these are the beaches busier times. The weekly data does have warning signs also, avoiding Sunnyside during the week, interestingly although this beach has the highest risk days it also pairs that with the safest days in the week.

\newpage
